---
title: "Weight Lifting Prediction"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
set.seed(1234)
```

# Introduction
****
The WLE(Weight Lifting Exercise) Dataset consists of data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.
Along with the data, it also contains a varaibe `classe` which denotes the quality of the exercie. The values `classe` can take are:

- **A**: exactly according to the specification
- **B**: throwing the elbows to the front
- **C**: lifting the dumbbell only halfway
- **D**: lowering the dumbbell only halfway
- **E**: throwing the hips to the front 

Our goal in this exercise is to predict the `classe` variable using the other variables in the dataset.

# Data
The data is available at the following URLs:

- [Training Data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)
- [Test Data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

We will train the model on the training data and then predict the `classe` variable for the test data.

## Downloading the data
Let us first download the data from the URLs.
```{r}
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

train_file <- "pml-training.csv"
test_file  <- "pml-testing.csv"

if (!file.exists(train_file)) {
  download.file(train_url, destfile=train_file)
}

if (!file.exists(test_file)) {
  download.file(test_url, destfile=test_file)
}
```

## Loading and Cleaning the Data
Now that we have downloaded the data, let us load it into R.
```{r cache=TRUE}
train_data <- read.csv(train_file)
test_data  <- read.csv(test_file)
```

Before working on the data, we should clean it. First, let us remove the columns we won't ned, like timestamps, user names, etc.
```{r}
columns_to_remove <- c(1:7)
train_data <- train_data[, -columns_to_remove]
test_data  <- test_data[, -columns_to_remove]
```

Next, let us remove the columns with missing values.
```{r}
non_na_cols <- colSums(is.na(train_data)) == 0 & colSums(is.na(test_data)) == 0
train_data <- train_data[, non_na_cols]
test_data  <- test_data[, non_na_cols]
```

Finally, let us set the `classe` variable as a factor.
```{r}
train_data$classe <- factor(train_data$classe)
```

## Splitting the Data
Since the test data does not have the `classe` variable, we will not be able to see how our model performs on it.
So, in order to estimate out of sample error, we will split the training data into two parts: training and validation.
We will train the model on the training data and then predict the `classe` variable for the validation data.
```{r}
inTrain <- createDataPartition(train_data$classe, p=0.8, list=FALSE)
training <- train_data[inTrain, ]
validation <- train_data[-inTrain, ]
```

# Methodology
We will use the **Random Forest** algorithm to predict the `classe` variable.
The parameters for our training algorithm will be:

- Methodology: Random Forests
- No. of folds for cross validation: 5
- No. of trees: 7
- Maximum depth of trees: 10

```{r cache=TRUE}
cv <- trainControl(method="cv", number=5)
model <- train(
    classe ~ ., 
    data=training, 
    method="rf", 
    trControl=cv, 
    ntree=7, 
    maxDepth=10
)
```

# Results
Now that we have trained the model, let us see how it performs on the training and validation data.

## Training Data
```{r}
train_pred <- predict(model, training)
conf_matrix <- confusionMatrix(train_pred, training$classe)
accuracy <- as.double(conf_matrix$overall["Accuracy"])
conf_matrix
```
The model performs really well on the training data with an accuracy of ``r accuracy*100`%`.

## Validation Data
```{r}
validation_pred <- predict(model, validation)
conf_matrix <- confusionMatrix(validation_pred, validation$classe)
accuracy <- as.double(conf_matrix$overall["Accuracy"])
conf_matrix
```
The model performs well on the validation data with an accuracy of ``r accuracy*100`%`.

# Predictions
Now that we have trained the model, let us predict the `classe` variable for the test data and save it to `test_predictions.csv`.
```{r}
test_pred <- predict(model, test_data)
test_pred_df <- data.frame(problem_id = test_data$problem_id, classe = test_pred)
write.csv(test_pred_df, file="test_predictions.csv", row.names=FALSE)
```

# Notes
- The code for this report is available at [this GitHub repository](https://github.com/InfinityAkshay/Practical-Machine-Learning)
- The report is generated using R Markdown
- The analysis is done with a seed of `1234`
- The packages used are:
    - `caret`: `r packageVersion("caret")`
    - `knitr`: `r packageVersion("knitr")`


